---
title: "Package stR"
author: "Alex Dokumentov"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package stR}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This vignette describes functionality and provides examples of how to use package stR.
stR implements method STR, and STR comes as an abbrevation of Seasonal-Trend decomposition by Regression and from R, a popular statistical software. The name is also reminiscent of STL, the method which inspired me for creation of STR.

## Introducation

There are many packages and methods which work with seasonal data. Thus the oldest methods for decomposition -- classical additive and multiplicative decompositions -- are implemented in package <code>stats</code>. Here is an example from the method's help:

```{r, fig.show='hold', fig.width = 7}
library(graphics)
m <- decompose(co2)
plot(m)
```


## Getting started

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## More complex examples

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## Even more Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
